\chapter{Introduction}
\label{cha:introduction}

As software becomes increasingly critical to modern infrastructure,
the proliferation of programming languages \citep{noauthor_tiobe_nodate}
has heightened the need for rigorous program verification.
It is essential to ensure that a program's execution aligns precisely
with the programmer's intent \citep{ringer_qed_2019}.
While this alignment is particularly vital when compilers translate
high-level code into machine-level representations,
the complexity of these translations introduces significant
risks of subtle behavioral shifts.

To address these challenges, the field of programming language semantics
provides the mathematical framework necessary to formalize the behavior of instructions.
Rather than viewing semantics as a modern invention,
it is more accurate to say that every programming language possesses
an underlying semantics, whether defined formally
or understood informally by developers.
By moving from an informal understanding to a rigorous mathematical
representation \citep{winskel_formal_1993},
we enable the use of formal methods to reason about program behavior
allowing verification of methods and programs for all valid inputs.

In academic settings, such as the course "Semantics of Programming Languages (1DL311)"
at Uppsala University, these definitions and proofs are typically presented
"on paper"~\citep{uppsala_syllabus_2023, uppsala_reading_2023}.
However, manual proofs are susceptible to human error~\citep{reliability_of_proof},
implicit assumptions, and logical gaps.
Furthermore, paper-based proofs frequently omit repetitive inductive steps,
relying on phrases like "the remaining cases are analogous"
or "it is easily seen",
shortcuts that are not permissible in a formal logic system~\citep{andersen_acceptable_2020}.

To achieve total precision, one can utilize \textit{mechanization}:
the process of implementing a formal system within a computer program,
such as a proof assistant, to verify every logical step automatically.
By mechanizing a proof, we introduce a layer of algorithmic validation
that ensures logical soundness and eliminates the ambiguity inherent in
hand-written derivations as was seen during the formalization of the Kepler conjecture~\citep{hales_formal_2017}.

This thesis investigates the process of translating pen-and-paper
semantic proofs into mechanized proofs using the
Lean proof assistant~\citep{theorem_proving_2026}.
The goal is to evaluate whether the techniques used in standard semantics
literature~\citep{uppsala_reading_2023} correspond directly to Lean's mechanisms,
or if the transition to machine-level precision requires a fundamental
reformulation of these proofs.


%
%
%
\label{intro:research_question}
\section{Research Question and Goals}
The primary objective of this work is to determine the "semantic gap"
between human-readable proofs and machine-checked logic.
Specifically, this thesis seeks to answer the following research questions:
\begin{itemize}
    \item \textbf{RQ1:} Do the proof techniques presented in the textbook~\citep{nielson_semantics_2007}
    map directly to Lean's internal mechanisms, or do they require significant
    structural reformulation?
    \item \textbf{RQ2:} What implicit assumptions or "pedagogical shortcuts"
    present in the literature must be made explicit to satisfy the Lean kernel?
\end{itemize}

%
%
%
\label{intro:motivation}
\section{Motivation}
\elias{Remove subsection and combine with 1.0} 
Informal proofs are designed for human consumption and often prioritize
intuition over exhaustive detail \hampus{this statement feels weired especially "humna consumption"}.
However, these omissions can hide underlying complexities or logical leaps \hampus{no proof of this} \missingcitation \todo{Cite a source discussing 'social processes' in proofs or the 'De Bruijn factor'}.
By identifying these points of divergence,
this research provides a roadmap for how educators can
structure pen-and-paper proofs to decrease ambiguity and better prepare students
for formal verification tasks.

%
%
%
\section{Scope}
\label{intro:scope}
This thesis will evaluate proofs from the course "Semantics of Programming Languages
(1DL311)" which uses the literature "Semantics with Applications: An Appetizer"~\citep{nielson_semantics_2007}.
will be limited to the evaluation to a few proofs from the chapter
"1. Introduction" ~\citep[p.~1]{nielson_semantics_2007} and
"2. Operational Semantics" ~\citep[p.~19]{nielson_semantics_2007}.
These chapters introduce the toy programming language "While", which will be
used in the proofs mechanized into Lean 4.

%
%
%
\section{Contribution}
\label{intro:contribution}
This thesis contribution to the field of programming language is the process of
mechanizing written programming language semantic proofs into Lean 4,
as well as which assumptions if any is needed to be stated explicitly in the proof
assistant that otherwise was implicitly defined in the written counterpart.

%
%
%
\label{intro:roadmap}
\section{Thesis Structure Roadmap} 
The journey of this thesis is organized into seven chapters,
progressing from theoretical foundations to the nuances of machine-checked proofs.
Following this introduction, Chapter 2 establishes the necessary background,
bridging the gap between the formal semantics of programming languages and the
technical mechanics of the Lean 4 proof assistant.
This section introduces the While language and explores how mathematical logic is
mapped onto Lean's inductive types and tactic-based proof state.
Chapter 3 then details the methodology employed,
describing the systematic approach used to translate traditional textbook
definitions into a mechanized framework.

The core technical contribution of this work is contained in Chapter 4,
which documents the formalization of both Natural Semantics (big-step)
and Structural Operational Semantics (small-step).
This chapter moves beyond simple definitions to prove critical properties
such as determinism and loop unrolling, culminating in a machine-verified
proof of equivalence between the two semantic styles when
applied to the \While language.
Having established this formal framework, Chapter 5 provides a critical
discussion of the findings.
It reflects on the "notation gap" between human-readable proofs
and computer-verified code, specifically addressing how the
complexity of the equivalence theorem was managed within Lean.

The final portion of the thesis situates these results within
the broader academic context.
Chapter 6 surveys related work, comparing the abstractions
used in this project with other modern formalization frameworks
and discussing the role of automation in engineering "trivial" proofs.
Finally, Chapter 7 offers concluding remarks, summarizing the primary
contributions of the research and identifying future pathways for the
mechanization of programming language theory.