%
%
%
%
%
%
\chapter{Background}
\label{cha:background}
\noindent
This chapter establishes the theoretical foundation necessary for the thesis.
We begin by specifying the \While language, which serves as the primary object
of study throughout this work. This provides the concrete syntax required to
explore the formal semantics of programming languages—the mathematical bridge
between a program's notation and its behavior. We then delve into Operational
Semantics, detailing both the Natural (big-step)~\citep[p.~20]{nielson_semantics_2007} and Structural (small-step)~\citep[p.~33]{nielson_semantics_2007}
approaches, followed by the primary proof techniques used in their analysis.
Finally, we introduce the Lean 4 proof assistant, detailing the mechanisms used
to translate these textbook definitions into machine-checked proofs~\citep{theorem_proving_2026}.

%
%
%
\section{Formal Semantics of Programming Languages}
\label{background:sec:formal_semantics_pl}
\
The formal semantics of a programming language is a rigorous mathematical
framework used to describe the behavior of programs \missingcitation \todo{Standard definition of formal semantics}. While a language's syntax
defines the rules for constructing valid programs (the "grammar"), its semantics
provides the "meaning"—specifying exactly how a computer should execute those
instructions \missingcitation \todo{Source for the syntax-vs-semantics distinction}.

\
In many industrial settings, the semantics of a language are described using
natural language manuals \missingcitation \todo{Provide example or source discussing the use of manuals/standards like C99 or Java Spec}. However, natural language is often prone to ambiguity,
leading to inconsistent implementations across different compilers or interpreters \missingcitation \todo{Cite a known study or famous example of ambiguity-related bugs (e.g., C undefined behavior)}.
Formal semantics eliminates this ambiguity by using mathematical structures—such
as sets, functions, and logic to define program execution \missingcitation \todo{Reference for the benefits of formal methods}.

By separating how an instruction is written (the syntax) such as
\begin{While_Code}
    if (x > 0) then {x = x} else {x = -x}
\end{While_Code} % \noindent
from the meaning (the semantics) such as "if x is less than 0 then negate the value, otherwise do nothing" it is possible to
create a deeper and more detailed understanding \hampus{this statement has no basis} \missingcitation \todo{Source discussing the benefits of abstraction in language design/analysis} of the programming language
and thereby the programs written.


\
Formal semantics are typically expressed through three main approaches:
Denotational (modeling programs as mathematical functions),
Axiomatic (using logic to define preconditions and postconditions),
and Operational \missingcitation \todo{Cite the seminal paper or textbook that categorizes these three (e.g., Scott/Strachey, Hoare, and Plotkin)}. This thesis focuses on the Operational approach, which describes
how a program transition from one state to another during execution \missingcitation \todo{Reference defining Operational Semantics specifically}.

In addition to these different semantics a representation of the language in question is needed.
This representation is built using five different components.
First of which is the so-called "\emph{Syntactic Categories}"

%
\subsection*{Syntactic Categories}
The syntactic categories of a language represent the alphabet.
This could include things such as number, variable arithmetic and boolean expressions.
It is also common to define so called "\emph{meta-variables}" which is used as a placeholder value for each category,
an example would be \(S\) which could represent a \emph{Statement} in a language, where \(S_1\) is another statement than \(S_2\) or \(S'\).

In addition to defining the categories and \emph{meta-variables}, it is important to define the grammar of the alphabet, effectively creating words out of the alphabet.
This is commonly done using BNF, short for \emph{Backus-Naur Form} \missingcitation \todo{Original BNF source or standard textbook definition}, an example can be seen in the \autoref{background:sec:while:syntactic}.
By providing the grammar it is possible to start building programs using the earlier defined alphabet, but this is just syntax,
we have yet to define meaning to our alphabet and words created from it.
To give the words meaning we use the next two components which is called "\emph{Semantic Domains}" and "\emph{Semantic Functions}"

%
\subsection*{Semantic Domains}
This component provides the semantic domains in which we will use to give our programs meaning \missingcitation \todo{Definition of Semantic Domains}.
A domain can be all natural numbers, or true and false, or even a custom domain such as program states or memory locations.
Next we need to know how the words in our alphabet is translated into our semantic domains, this is done by the "\emph{Semantic Functions}" mentioned earlier.

%
\subsection*{Semantic Functions}
This component provides the translation from our words to our domains.
It is common to define a custom notation to mark when a semantic function is being applied,
as it is common to use Natural numbers such as \(1\), \(2\), \(3\), etc., both as words in our language but also as the result of a semantic function.

Whilst this component describes a functional definition, the actual function definitions are defined in the next component which is called "\emph{Semantic Equations}".

%
\subsection*{Semantic Equations}
This component is responsible for defining on a case by case basis how exactly the semantic function translates the words given by our grammar into our semantic domains.
Often this is done compositionally \missingcitation \todo{Source for the Principle of Compositionality in semantics}, which means that we might define a multiplication as multiplying the result of two recursive calls on its components.
Lastly we define the last component known as "\emph{Auxiliary Components}".

\subsection*{Auxiliary Components}
This is the last and final component used to create a representation of a programming language.
Auxiliary components is a broad way of saying "everything else" this could include the definition of what a program state is,
or memory locations or even environments for dynamic/static function linking.
This section might contain functions and definitions that could occur in a previous component in a different paper,
this occurs as a result from picking which approach to use, and what the main focus is.


%
%
%
\section{The \While language}
\label{background:sec:while}
\ Programming language semantics needs a language to study, and commonly used
programming languages are often a bad fit due to the constant
evolution of languages, operating systems, and hardware \missingcitation \todo{A source arguing for the use of "toy" languages for formal analysis (e.g., Nielson)}. \
To address this, a common 
praxis is to invent a "toy" programming language that lacks a physical implementation 
but possesses rigorous rules \missingcitation \todo{Source on the educational/formal value of toy languages}. In this case, we use \While, a simplified imperative language 
defined in the literature of our case study ~\citep[p.~7]{nielson_semantics_2007}.

Like most programming languages, \While is built using statements and expressions. \
Informally, the language supports the following statements:
\begin{itemize}
    \item assignment,
    \item skip (no effect),
    \item if-then-else,
    \item while-loop,
    \item sequential composition.
\end{itemize}
This allows us to build simple programs where we use the sequential statement to connect
two operations, such as "assign $x$ to 4, then skip." Before we can define how these 
programs behave, we must first define the formal building blocks of the language's 
structure: its \emph{Syntactic Categories}.

%
%
\subsection*{Syntactic Categories}
\label{background:sec:while:syntactic}
The \While language is constructed from several basic syntactic sets. These sets 
define the "alphabet" and "words" available to the programmer:
\begin{itemize}
    \item numbers \Num, which consists of positive integers including 0,
    \item variables \Var, which consist of a letter followed by letters or numbers,
    \item arithmetic expressions \Aexp,
    \item boolean expressions \Bexp,
    \item statements \Stm.
\end{itemize}

To manipulate these sets in formal proofs, we introduce \emph{meta-variables}. 
By convention, we let specific letters range over these sets so that $n$ always 
implies a numeral, and $S$ always implies a statement:
\[
\begin{array}{ll}
    n  & \mbox{ranges over numerals, } \Num,\\[0.6em]
    x  & \mbox{ranges over variables, } \Var,\\[0.6em]
    a  & \mbox{ranges over arithmetic expressions, } \Aexp,\\[0.6em]
    b  & \mbox{ranges over boolean expressions, } \Bexp,\\[0.6em]
    S  & \mbox{ranges over statements, } \Stm.
\end{array}
\]
Each variable can be primed or sub-scripted (e.g., \(n', n_0, n_{317}\)) to create 
distinct instances that still belong to the same set. Using these meta-variables, 
we define the formation rules using Backus-Naur Form (BNF) notation \missingcitation \todo{Reference for the standard BNF notation used here}:
\[
\begin{array}{lcl}
    n &::=& 0 \mid 1 \mid n'0 \mid n'1 \\[0.6em]
    x &::=& [a-zA-Z]([a-zA-Z] \mid [0-9])^{*} \\[0.6em]
    a &::=& n \mid x \mid a_1 + a_2 \mid a_1 \star a_2 \mid a_1 - a_2 \\[0.6em]
    b &::=& \mathbf{true} \mid \mathbf{false} \mid a_1 = a_2 \mid a_1 \leq a_2 \mid \neg b' \mid b_1 \land b_2 \\[0.6em]
    S &::=& x := a \mid \mathbf{skip} \mid S_1; S_2 \mid \mathbf{if}\ b\ \mathbf{then}\ S_1\ \mathbf{else}\ S_2 \mid \mathbf{while}\ b\ \mathbf{then}\ S'
\end{array}
\]
The symbol "\(::=\)" denotes that a category "can be" a specific form, while "\(\mid\)" 
denotes an alternative. This is an \emph{inductive definition} \missingcitation \todo{Source explaining inductive definitions of syntax}: for example, 
an arithmetic expression $a$ can be a simple number $n$, or it can be a 
compound expression like $a_1 + a_2$, where $a_1$ and $a_2$ are themselves 
already formed arithmetic expressions.

With the syntax established, we now need a mathematical "universe" to map these 
symbols into. This is provided by the \emph{Semantic Domains}.

%
%
\subsection*{Semantic Domains}
While the syntax defines what we can write, the semantic domains define what those 
writings actually \emph{mean} in terms of mathematical objects \missingcitation \todo{Definition of Semantic Domains in this context}. We map \While 
syntax into the following sets:
\begin{itemize}
    \item Truth values \(\Truth = \{ \mathbf{tt}, \mathbf{ff} \}\), where \textbf{tt} represents mathematical truth.
    \item Integers \(\mathbb{Z}\), the set $\{ \dots, -1, 0, 1, \dots \}$.
    \item Program states \State, representing the memory configuration.
\end{itemize}
We use the meta-variable $s$ to range over states. To bridge the gap between 
the syntax of \Aexp or \Bexp and these mathematical domains, 
we require \emph{Semantic Functions} \missingcitation \todo{Standard reference for Semantic Functions}.

%
%
\subsection*{Semantic Functions}
Semantic functions serve as the "translators" from syntax to semantics. We 
standardize this translation using 'semantic brackets' \([\![\cdot]\!]\) to 
enclose the syntax being evaluated \missingcitation \todo{Reference for the Denotational style "Oxford Brackets" notation}. This separates the \While language from 
the mathematical "meta-language" used to describe it. To distinguish between 
different types of evaluation, we prefix the brackets with calligraphic letters:
\[
\begin{array}{ll}
    \mathcal{N} : \mathbf{Num} \to \mathbb{Z} & \mbox{(Translating numerals to integers)} \\[0.6em]
    \mathcal{A} : \mathbf{Aexp} \to (\mathbf{State} \to \mathbb{Z}) & \mbox{(Evaluating arithmetic in a state)} \\[0.6em]
    \mathcal{B} : \mathbf{Bexp} \to (\mathbf{State} \to \mathbf{T}) & \mbox{(Evaluating booleans in a state)}
\end{array}
\]
In the definitions for $\mathcal{A}$ and $\mathcal{B}$, the function takes a piece 
of syntax and returns a \emph{function} that expects a state $s$. We often 
write $\mathcal{A}[\![a]\!]s$ to mean "the integer result of expression $a$ 
evaluated in state $s$."

%
%
\subsection*{Semantic Equations}
The equations below provide the rules for our semantic functions. These are 
defined \emph{compositionally}, meaning the value of a complex expression depends 
solely on the values of its sub-expressions \missingcitation \todo{Formal definition of compositional semantics}. 

For numerals, we define the translation of binary-style strings to integers:
\[
\begin{array}{rcl}
    \mathcal{N}[\![0]\!] &=& 0 \\[0.6em]
    \mathcal{N}[\![1]\!] &=& 1 \\[0.6em]
    \mathcal{N}[\![n 0]\!] &=& 2 \cdot \mathcal{N}[\![n]\!] \\[0.6em]
    \mathcal{N}[\![n 1]\!] &=& 2 \cdot \mathcal{N}[\![n]\!] + 1
\end{array}
\]
Arithmetic expressions $\mathcal{A}$ are evaluated by looking up variables in 
the state $s$ or performing mathematical operations ($+, \cdot, -$) on the 
results of sub-expressions:
\[
\begin{array}{rcl}
    \mathcal{A}[\![n]\!]s &=& \mathcal{N}[\![n]\!] \\[0.6em]
    \mathcal{A}[\![x]\!]s &=& s \, x \\[0.6em]
    \mathcal{A}[\![a_1 + a_2]\!]s &=& \mathcal{A}[\![a_1]\!]s + \mathcal{A}[\![a_2]\!]s \\[0.6em]
    \mathcal{A}[\![a_1 \star a_2]\!]s &=& \mathcal{A}[\![a_1]\!]s \cdot \mathcal{A}[\![a_2]\!]s \\[0.6em]
    \mathcal{A}[\![a_1 - a_2]\!]s &=& \mathcal{A}[\![a_1]\!]s - \mathcal{A}[\![a_2]\!]s
\end{array}
\]
Boolean expressions $\mathcal{B}$ follow a similar pattern, returning \textbf{tt} 
or \textbf{ff} based on the mathematical truth of the comparisons:
\[
\begin{array}{rcl}
    \mathcal{B}[\![\mathbf{true}]\!]s &=& \mathbf{tt} \\[0.6em]
    \mathcal{B}[\![\mathbf{false}]\!]s &=& \mathbf{ff} \\[0.6em]
    \mathcal{B}[\![a_1 = a_2]\!]s &=& \left\{
        \begin{array}{@{}l@{\quad}l@{}}
            \mathbf{tt} & \mbox{if } \mathcal{A}[\![a_1]\!]s = \mathcal{A}[\![a_2]\!]s \\
            \mathbf{ff} & \mbox{if } \mathcal{A}[\![a_1]\!]s \not = \mathcal{A}[\![a_2]\!]s
        \end{array}
    \right. \\[1.2em]
    \mathcal{B}[\![a_1 \leq a_2]\!]s &=& \left\{
        \begin{array}{@{}l@{\quad}l@{}}
            \mathbf{tt} & \mbox{if } \mathcal{A}[\![a_1]\!]s \leq \mathcal{A}[\![a_2]\!]s \\
            \mathbf{ff} & \mbox{if } \mathcal{A}[\![a_1]\!]s > \mathcal{A}[\![a_2]\!]s
        \end{array}
    \right. \\[1.2em]
    \mathcal{B}[\![\neg b]\!]s &=& \left\{
        \begin{array}{@{}l@{\quad}l@{}}
            \mathbf{tt} & \mbox{if } \mathcal{B}[\![b]\!]s = \mathbf{ff} \\
            \mathbf{ff} & \mbox{if } \mathcal{B}[\![b]\!]s = \mathbf{tt}
        \end{array}
    \right. \\[1.2em]
    \mathcal{B}[\![b_1 \land b_2]\!]s &=& \left\{
        \begin{array}{@{}l@{\quad}l@{}}
            \mathbf{tt} & \mbox{if } \mathcal{B}[\![b_1]\!]s = \mathbf{tt} \mbox{ and } \mathcal{B}[\![b_2]\!]s = \mathbf{tt} \\
            \mathbf{ff} & \mbox{if } \mathcal{B}[\![b_1]\!]s = \mathbf{ff} \mbox{ or } \mathcal{B}[\![b_2]\!]s = \mathbf{ff}
        \end{array}
    \right.
\end{array}
\]

Throughout these definitions, we have relied on the notion of a "state" to store 
variable values. We conclude by formally defining this context.

%
%
\subsection*{Auxiliary Components}
The \emph{State} is the primary auxiliary component of our semantics. Formally, 
a state is a function that maps each variable to its current integer value:
\[
s : \mathbf{Var} \to \mathbb{Z}
\]
In this language, we assume the state is a \emph{total function}; if a variable 
has not been explicitly assigned, it is mapped to 0 by default. 

To describe how statements change the state (such as in an assignment $x := a$), 
we use the notation $s[x \mapsto v]$ to represent a new state that is identical 
to $s$ except that the variable $x$ now maps to the value $v$:
\[
s[x \mapsto v]y = \left\{
    \begin{array}{ll}
        v & \mbox{if } y = x \\
        s \, y & \mbox{if } y \neq x
    \end{array}
\right.
\]
This "substitution" notation is fundamental to describing the execution of 
imperative commands in \While~\citep[p.~13]{nielson_semantics_2007}.

The syntactic categories and semantic functions defined above provide the necessary
vocabulary for our toy language. Yet, the question remains: how do we formally
describe the step-by-step transition of a program from one configuration to another?
To answer this, we move from the definition of the language itself to the study
of Formal Semantics of Programming Languages, which in this case
will take focus on Operational Semantics.


%
%
%
\section{Operational Semantics}
\todo{Brief What?} \noindent
Operational semantics describes the meaning of a program by specifying how it
executes on an abstract machine \missingcitation \todo{Reference for the abstract machine definition of OS}. Unlike other semantic approaches, it is
"execution-oriented," focusing on the transitions between program states.
A state transition is typically represented as a relation between a
configuration—consisting of a statement and the current state—and a resulting
state or a subsequent configuration \missingcitation \todo{Reference for transition relations/configurations}.
\todo{Show state transition notation and describe it, lean on The Formal Semantics of Programming Languages: An Introduction}

In this work, we examine two primary styles of operational semantics:
\emph{Natural Semantics} (Big-Step) and
\emph{Structural Operational Semantics} (Small-Step).

\subsection*{Natural Operational Semantics (Big-Step)}
\label{background:os:ns}
\todo{Breif introduction to NS, reader will have read a course on semantics,
this section is here as a reminader to the reader.}
Natural Operational Semantics, often shortened to just Natural Semantics (NS), or even big-step semantics
is the view that a given Statement \(S\) and a starting state \(s\) can evaluate into a finalized state \(s'\).
This view is that each operation runs to completion.

As with most semantics we define deduction rules to formulate how the statements are evaluated.
In the textbook we are given the following table of rules

\hampus{Show Comp NS rules for \While describing them}
\begin{table}[ht]
    \[
    \begin{array}{ll}
        \mbox{[ass}_{ns}\mbox{]} &
\begin{prooftree}
  \hypo{} % Empty hypothesis for an axiom
  \infer1{
    \langle x := a, s \rangle \to s[x \mapsto \mathcal{A}[\![a]\!] s]
    }
\end{prooftree} \\ \\

        \mbox{[skip}_{ns}\mbox{]} &
\begin{prooftree}
  \hypo{} % Empty hypothesis for an axiom
  \infer1{
    \langle \mathtt{skip}, s \rangle \to s
    }
\end{prooftree} \\ \\

        \mbox{[comp}_{ns}\mbox{]} &
\begin{prooftree}
  \hypo{\langle S_1, s \rangle \to s'}
  \hypo{\langle S_2, s \rangle \to s''}
  \infer2{
    \langle S_1;\ S_2, s \rangle \to s''
    }
\end{prooftree} \\ \\

        \mbox{[if}_{ns}^{tt}\mbox{]} &
\begin{prooftree}
  \hypo{\langle S_1, s \rangle \to s'}
  \infer1[
    if $\mathcal{B}[\![b]\!]s = \mathbf{tt}$
    ]{
    \langle \mathtt{if}\ b\ \mathtt{then}\ S_1\ \mathtt{else}\ S_2, s \rangle \to s'
    }
\end{prooftree} \\ \\

        \mbox{[if}_{ns}^{ff}\mbox{]} &
\begin{prooftree}
  \hypo{\langle S_2, s \rangle \to s'}
  \infer1[
    if $\mathcal{B}[\![b]\!]s = \mathbf{ff}$
    ]{
    \langle \mathtt{if}\ b\ \mathtt{then}\ S_1\ \mathtt{else}\ S_2, s \rangle \to s'
    }
\end{prooftree} \\ \\

        \mbox{[while}_{ns}^{tt}\mbox{]} &
\begin{prooftree}
  \hypo{\langle S, s \rangle \to s'}
  \hypo{\langle \mathtt{while}\ b\ \mathtt{do}\ S, s' \rangle \to s''}
  \infer2[
    if $\mathcal{B}[\![b]\!]s = \mathbf{tt}$
    ]{
    \langle \mathtt{while}\ b\ \mathtt{do}\ S, s \rangle \to s''
    }
\end{prooftree} \\ \\

        \mbox{[while}_{ns}^{ff}\mbox{]} &
\begin{prooftree}
  \hypo{} % Empty hypothesis for an axiom
  \infer1[
    if $\mathcal{B}[\![b]\!]s = \mathbf{ff}$
    ]{
    \langle \mathtt{while}\ b\ \mathtt{do}\ S, s \rangle \to s
    }
\end{prooftree}

    \end{array}
    \]
\end{table}

The first rule is an axiom, a rule that can be used regardless of the premises at hand,
this is a fact which is always true in this language. This should not be confused with the fact
that you can use it wherever, as an assignment cant be used by itself to prove that we can build a statement from two smaller statements.

An axiom is marked by the empty space above the line where the premises are needed,
The assignment rule uses the state update syntax defined for the while language to show that the Statement \(x := a\) in the state \(s\)
will result in the final state \(s[x \mapsto \mathcal{A}[\![a]\!]s]\).

In that same regard we can see that skip is also an axiom, but in this case the evaluation of the \(\mathtt{skip}\) statement does not modify the state \(s\)

The next rule in the table is the composition rule, which states that given two statement \(S_1\) and \(S_2\) of some shape, that terminates into the final states
\(s'\) from the initial state \(s\) and \(s''\) from the initial state \(s'\) respectively, this means we can take the output \(s'\) from \(S_1\),
and feed it into \(S_2\) to get \(s''\).
Compared to the assignment and the skip rules, we must prove that \(\langle S_1, s \rangle \to s'\) and \(\langle S_2, s \rangle \to s''\) before we can use it.

The next rule is the [if$_{ns}^{tt}$] rule, where the concept of side-condition is used.
A side condition is much like a premise in that we need to prove that the side condition is true before we use the rule, but compared to the premises, we do not
need to create this proof directly in the deduction tree we are building, but instead can provide an external proof for why the condition is true.
This removes this proof tree from the whole allowing the textbook to focus on the important parts, whilst leaving the more trivial proofs outside the deduction trees.

The remaining rules are straight forward combination of premises and side conditions to formulate how to build
conditional statements when the condition is false, as well as while-loops for both when condition is true and when it is false.

Yet again notice that we can only formulate deduction trees, or so-called transitions given these rules, this means that things that never terminates into a final state,
can not be proven using these rules, an example would be the statement
\[
    \mathtt{while}\ \mathbf{true}\ \mathtt{do}\ \mathtt{skip}
\]
which would
create an infinite growing deduction tree due to the second premise \(\langle \mathtt{while}\ b\ \mathtt{do}\ S, s' \rangle \to s''\) is exactly the same thing we need to prove
for the final conclusion, meaning that the proof becomes infinitely recursive. This can also occur for statements that always modifies the state, but never reaches a final state.


%
%
%
\subsection*{Structural Operational Semantics (Small-Step)}
\label{background:os:sos}
\todo{Breif introduction to SOS, reader will have read a course on semantics,
this section is here as a reminader to the reader.}
While Natural Semantics describes the overall result of execution, Structural Operational Semantics (SOS), or "Small-Step" semantics,
describes the individual steps of the computation.
This approach is particularly useful for reasoning about interleaved execution,
non-termination, and program intermediate states.In SOS, we distinguish between
two types of configurations:
\begin{itemize}
    \item Intermediate configurations: $\langle S, s \rangle$,
representing that the statement $S$ is to be executed from state $s$.

    \item Terminal configurations: $s$, representing that the program has terminated in state $s$.
\end{itemize}
The transition relation is denoted by $\Rightarrow$. 
A transition can either lead to a new intermediate configuration,
$\langle S, s \rangle \Rightarrow \langle S', s' \rangle$ (representing one step of execution),
or directly to a terminal state, $\langle S, s \rangle \Rightarrow s'$.

Comparison of Transition Styles
Unlike Natural Semantics,
where the derivation tree for a complex statement is constructed all at once,
an SOS derivation represents a single "pulse" of the abstract machine. This can be shown in the rules
\begin{table}[ht]
    \[
    \begin{array}{ll}
        \mbox{[ass}_{sos}\mbox{]} &
\begin{prooftree}
  \hypo{} % Empty hypothesis for an axiom
  \infer1{
    \langle x := a, s \rangle \Rightarrow s[x \mapsto \mathcal{A}[\![a]\!] s]
    }
\end{prooftree} \\ \\

        \mbox{[skip}_{sos}\mbox{]} &
\begin{prooftree}
  \hypo{} % Empty hypothesis for an axiom
  \infer1{
    \langle \mathtt{skip}, s \rangle \Rightarrow s
    }
\end{prooftree} \\ \\
        \mbox{[comp}_{sos}^1\mbox{]} &
\begin{prooftree}
\hypo{\langle S_1, s \rangle \Rightarrow \langle S_1', s' \rangle}
\infer1{
    \langle S_1; S_2, s \rangle \Rightarrow \langle S_1'; S_2, s' \rangle
    }
\end{prooftree} \\ \\

\mbox{[comp}_{sos}^2\mbox{]} &
\begin{prooftree}
    \hypo{\langle S_1, s \rangle \Rightarrow s'}
    \infer1{
        \langle S_1; S_2, s \rangle \Rightarrow \langle S_2, s' \rangle}
\end{prooftree} \\ \\

        \mbox{[if}_{sos}^{tt}\mbox{]} &
\begin{prooftree}
  \hypo{}
  \infer1[
    if $\mathcal{B}[\![b]\!]s = \mathbf{tt}$
    ]{
    \langle \mathtt{if}\ b\ \mathtt{then}\ S_1\ \mathtt{else}\ S_2, s \rangle \Rightarrow \langle S_1, s \rangle
    }
\end{prooftree} \\ \\

        \mbox{[if}_{sos}^{ff}\mbox{]} &
\begin{prooftree}
  \hypo{}
  \infer1[
    if $\mathcal{B}[\![b]\!]s = \mathbf{ff}$
    ]{
    \langle \mathtt{if}\ b\ \mathtt{then}\ S_1\ \mathtt{else}\ S_2, s \rangle \Rightarrow \langle S_2, s \rangle
    }
\end{prooftree} \\ \\

        \mbox{[while}_{sos}\mbox{]} &
\begin{prooftree}
  \hypo{}
  \infer1{
    \langle \mathtt{while}\ b\ \mathtt{do}\ S, s \rangle \Rightarrow \langle \mathtt{if}\ b\ \mathtt{then}\ (S;\ \mathtt{while}\ b\ \mathtt{do}\ S)\ \mathtt{else}\ \mathtt{skip}, s \rangle
    }
\end{prooftree}
\end{array}
    \]
\end{table}
Here we can see that almost all rules except for the compositional rules of which there is now two of, are axioms.
Where assign, and skip both still terminates after a single pulse/operation, the other rules such as the if conditions and the singular while rule
simply unwraps the conditional part and loop.
A true conditional is simply just the true branch, and vice versa for the false case.
A while-loop is either a single execution of the statement followed by the loop again, or nothing depending on the condition.


To find the final state of a program,
one must look at the transitive closure of the transition relation, denoted $\Rightarrow^*$.
The rules for SOS are structured to "reduce" the statement piece by piece.
For example, the composition rule $[comp_{sos}^1]$ allows us to execute the first step of
$S_1$ while keeping $S_2$ "on the stack":

Handling Non-termination
In Natural Semantics, a looping program like \\
$\mathtt{while}\ \mathbf{true}\ \mathtt{do}\ \mathtt{skip}$ simply
has no valid derivation tree. In SOS, this same program produces an infinite sequence of configurations:
$$
\langle \mathtt{while}\ \mathbf{true}\ \mathtt{do}\ \mathtt{skip}, s \rangle \Rightarrow \dots \Rightarrow \langle \mathtt{while}\ \mathbf{true}\ \mathtt{do}\ \mathtt{skip}, s \rangle \Rightarrow \dots
$$
This is still not a valid program in SOS, but it allows us to distinguish between a program that is "stuck" (no transition possible)
and one that is "diverging" (infinite transitions possible).



%
%
%
\section{The Lean 4 Proof Assistant}
\label{background:sec:lean_4}
\hampus{Show by example, pen-and-paper proof proves evenness by induction, this is done like this in lean \dots}
Lean 4 is a functional programming language and an interactive theorem prover
based on \textit{Dependent Type Theory} \citep{theorem_proving_2026}.
For the readers familiar with first- and second-order logic,
Lean can be viewed as a rigorous environment for defining mathematical objects
and constructing formal proofs that are mechanically verified by a small logical kernel.

\subsection{Defining Syntax: Inductive Types}
In the textbook by \cite{nielson_semantics_2007},
the syntax of a programming language is often defined using Backus-Naur Form (BNF).
In Lean, we represent these syntactic categories using \texttt{inductive} types.
An inductive type defines a set by listing its constructors.
For example, the syntactic category of arithmetic expressions (\textit{Aexp}) is defined as:

\begin{Lean_Code}
inductive Aexp where
  | num (n : Num)          -- Constants
  | var (x : Var)          -- Variables
  | add (a1 a2 : Aexp)     -- Addition
\end{Lean_Code}
\noindent
This maps directly to the mathematical notion of a set defined by structural induction.
Each constructor (e.g., \texttt{add}) is a function
that takes existing elements of the set and produces a new one.

\subsection{Defining Semantics: Relations as Propositions}
A core challenge in this thesis is mapping the "Big-Step" operational semantics
found in literature to Lean's internal logic. In a traditional math course,
the evaluation of a statement is written as a relation: $\langle S, s \rangle \to_{ns} s'$. 

In Lean, this relation is defined as an \texttt{inductive} type that returns
a \texttt{Prop} (Proposition).
This is a crucial distinction:
\begin{itemize}
    \item \textbf{Data Types} (like \texttt{Aexp}) are used to build programs.
    \item \textbf{Propositions} (like \texttt{big\_step}) are used to make claims about those programs.
\end{itemize}
\noindent
When we define \texttt{big\_step} inductively, each constructor represents an
\textbf{inference rule}. For instance, the rule for statement composition (\texttt{comp}) in Lean is:

\begin{Lean_Code}
| comp (h_left : §\(\langle\)§S1, s§\(\rangle\)§ §\(\to_{ns}\)§ s') (h_right : §\(\langle\)§S2, s'§\(\rangle\)§ §\(\to_{ns}\)§ s'') :
    §\(\langle\)§S1; S2, s§\(\rangle\)§ §\(\to_{ns}\)§ s''
\end{Lean_Code}
This code defines a rule of inference. To a reader familiar with natural deduction,
this is a direct mapping of the standard horizontal-bar notation:
\[
\frac{\langle S_1, s \rangle \to_{ns} s' \quad \langle S_2, s' \rangle \to_{ns} s''}{\langle S_1; S_2, s \rangle \to_{ns} s''}
\]
The names \texttt{h1} and \texttt{h2} represent the \textit{hypotheses} or \textit{premises}
required to reach the conclusion below the colon.

\subsection{The Proof State and Tactics}
Unlike a prose proof, which is a static piece of text,
a proof in Lean is constructed using \textbf{tactics}.
A tactic is a command that transforms a "goal" (the statement to be proven) into simpler sub-goals.

When reading the proofs in this thesis,
it is helpful to understand the workflow:
Lean maintains a \textit{proof state} consisting of the current \textbf{context}
(known facts and variables) and the \textbf{goal}.
A tactic like \texttt{intro} moves an assumption into the context,
while \texttt{cases} performs a proof by exhaustion,
forcing Lean to consider every possible way a specific relation could have been formed.

\subsection{Mapping Lean to Mathematical Logic}
The following table provides a quick reference for mapping Lean's core keywords
to their respective functionalities in standard mathematical and logical discourse.

\begin{table}[h]
\centering
\begin{tabular}{llp{7cm}}
\hline
\textbf{Lean Keyword} & \textbf{Logical Equivalent} & \textbf{Functionality} \\ \hline
\texttt{inductive}    & Inductive Definition      & Defines a set or a relation via a list of rules/constructors. \\
\texttt{Prop}         & Proposition               & The type of statements that have truth values. \\
\texttt{theorem}      & Theorem / Lemma           & A claim that requires a formal proof term. \\
\texttt{intro}        & Implication Introduction  & "Assume $P$ is true" or "Let $x$ be an arbitrary element." \\
\texttt{exact}        & Q.E.D.                    & Completes a goal by providing the required evidence directly. \\
\texttt{cases}        & Proof by Exhaustion       & Deconstructs an object or relation into its possible forms. \\
\texttt{apply}        & Modus Ponens              & Uses a known rule or theorem to satisfy the current goal. \\
\texttt{rfl}          & Reflexivity               & Proves an equality of the form $A = A$ by definition. \\ \hline
\end{tabular}
\caption{Mapping of Lean 4 keywords to mathematical concepts.}
\end{table}

\subsection{Notation Conventions}
To bridge the gap between code and literature, this thesis employs custom notation.
We use semantic brackets $\mathcal{A}[\![ a ]\!] s$ to denote the evaluation of
an expression $a$ in state $s$, and the notation $s[x \mapsto z]$ to denote a
state update where variable $x$ is assigned value $z$. These notations are not
merely for display; they are mapped directly to Lean functions through the
\texttt{notation} and \texttt{unexpander} mechanisms, ensuring that the mechanized
proof code remains readable to a human audience.
