

%
%
%
%
%
%
\chapter{Method}

This chapter outlines the methodology used to mechanize the semantic proofs
from the textbook~\citep{nielson_semantics_2007}, ensuring a systematic and traceable translation from
pen-and-paper arguments to formal verification in Lean 4~\citep{theorem_proving_2026}.
The core of this
thesis lies in not just proving the theorems, but in analyzing the transformation
process itself. The method is designed to expose the implicit assumptions and
reasoning patterns that must be made explicit during formalization.

The methodology can be broken down into four main stages:

\begin{enumerate}
    \item \textbf{Mechanization of Proofs:} The first step involves translating the proofs
    from the literature into Lean 4. A key principle here is to mirror the original proof
    strategy as closely as possible\missingcitation\todo{Optional: Mention if this "alignment" approach follows a specific formalization philosophy, like De Bruijn’s or similar comparative studies.}.
    When the textbook specifies "induction on the shape
    of the derivation tree," the \texttt{induction} tactic is used. Similarly, when a proof
    proceeds by "case analysis," the \texttt{cases} tactic is employed. This ensures that
    the resulting formal proof maintains a clear correspondence with the original argument.

    \item \textbf{Side-by-Side Comparison:} Once a proof is successfully mechanized,
    it is presented alongside its pen-and-paper counterpart. This direct comparison,
    as seen in Chapter \ref{formal_ns_sos_equiv}, serves to ground the formalization
    and provides a clear basis for identifying the divergences and additional machinery
    required by the mechanized approach.

    \item \textbf{Evaluation of Transformations:} With both versions of the proof available,
    a detailed evaluation is conducted to identify the specific changes required to make the
    formal proof work. This involves pinpointing where the mechanization required additional
    steps, such as:
    \begin{itemize}
        \item Explicitly handling cases that lead to contradictions.
        
        \item Introducing necessary assumptions or auxiliary lemmas that were implicit in
        the original text.
        
        \item Generalizing proof statements to facilitate induction.
        
        \item Formalizing concepts that the literature took for granted
        (e.g., multi-step relations).
    \end{itemize}

    \item \textbf{Summarization of Differences:} Finally, the findings from the evaluation
    are synthesized into a comprehensive summary of the differences between informal and
    formal reasoning. This summary, presented in the discussion chapter, distills the
    recurring patterns of transformation and articulates the core principles of mechanization
    that emerge from the case studies.
\end{enumerate}

This structured approach ensures that the thesis provides not only a set of verified
theorems but also a reflective analysis of what it means to translate informal mathematical
reasoning into a fully formal, machine-checked system\missingcitation\todo{Cite a source that discusses the general challenge of 'The Formalization Gap' or 'Informal vs Formal proofs' to justify why this reflection is valuable.}.
The next chapter will present the
results of this methodology, beginning with the side-by-side comparison of the
pen-and-paper and mechanized proofs for the inductive types.