

%
%
%
%
%
%
\chapter{Related Work}
The mechanization of programming language semantics requires navigating a
trade-off between the high-level intuition of pedagogical materials and the
low-level rigor of a proof assistant. This research is situated among several
key efforts to bridge this "semantic gap."

%
%
%
\section{Educational Abstractions vs. General Purpose Provers}
To lower the entry barrier for students, tools like
SASyLF by~\cite{SASyLF} provide a specialized environment where proofs are constructed in a
"math-like" syntax. SASyLF~\citep{SASyLF}
prioritizes the student's mental model by
abstracting away much of the "syntactic noise" identified in the different proofs.
In contrast, this thesis uses Lean 4, a general-purpose theorem prover.

Using Lean 4 lets us make the trade-offs explicit: we lose some of the
``math-like'' surface syntax and automated scaffolding that educational tools hide,
but we gain a fully general, reusable proof infrastructure and stronger guarantees
about what is actually proved. Concretely, we analyze how the ``pedagogical
shortcuts'' listed in RQ2—implicit side conditions, compressed notation, and
elided intermediate steps—must be reconstructed or made explicit in a
general-purpose prover.

%
%
%
\section{The Variable Binding Problem and Scoping}
The difficulty of formalizing
variable binding, the "infrastructure" of names and scopes is the central
theme of the POPLMark Challenge~\citep{aydemir_mechanized_2005}. Traditional solutions often involve complex
techniques, such as Nominal Techniques in Isabelle/HOL~\citep{urban_nominal_Isabelle_2008}, which automate the
handling of $\alpha$-equivalence and variable freshness. However, as noted in
the analysis of the textbook~\citep{nielson_semantics_2007}, the target language assumes a Global
Scope with no variable shadowing or undefined variables. Consequently, this
work intentionally bypasses the binding complexities discussed in the POPLMark~\citep{aydemir_mechanized_2005} 
literature. This allows for a focused evaluation of other "verbatim" translation
issues, such as:
The transition from intuitive side conditions to explicit premises.
The requirement for decidable equality on variable strings among others.

%
%
%
\section{Engineering the "Trivial": Scaffolding and Tactics}
Even in a language with simple scope,
the engineering effort required to satisfy a mechanized
proof assistant remains high.

This challenge is underscored by QED at Large~\citep{ringer_qed_2019},
a survey of formally verified software which highlights that
"proof engineering", specifically the creation of scaffolding
lemmas, is often the most time-consuming part of formalization.
This insight directly aligns with the finding that textbook logic
for state updates ($s[x \to v]$) requires multiple "hidden"
lemmas in Lean to prove identity and extensionality.

On a larger scale, the formalization of the C standard,
such as the CH2O project~\citep{mechanizedCSpecification},
further demonstrates how informal prose must be meticulously
converted into a rigid, constructive logic.

By mirroring this extensive process on a pedagogical scale,
this thesis investigates how textbook assumptions,
which are often taken for granted, must be made entirely explicit.

%
%
%
\section{Modern Formalization Frameworks}
The recent Modular Lean 4 Framework for Lambda
Calculi~\citep{ramos2025modularLean4framework} provides a look at how Lean 4 can be used to build highly structured,
re-usable metatheory. While that framework focuses on the complexity of products,
sums, and normalization, this work focuses on fidelity to the curriculum. While a
modular framework re-engineers the logic for efficiency in the prover, this "verbatim"
approach tests whether Lean can be used as a direct mirror for existing educational
literature.

%
%
%
\section{Automation and the Poincaré Principle}
Finally, the gap in reasoning
about numbers is bridged in this work using Lean's \texttt{linarith} tactic.
This follows the "Poincaré Principle"~\citep{barendregt_autarkic_2002}, the idea that trivial computational steps
should be handled by the machine rather than the user. By utilizing \texttt{linarith},
we ensure that the focus of the formalization remains on the semantics of the
language rather than the intricacies of natural number induction, which is not
the primary focus of the textbook~\citep{nielson_semantics_2007}.